### 2013-01-31 trunk/htdocs/surfactant.tex
Revamped documentation
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1902 40876441-3496-0410-a613-a2fca35b2ae2

@@

@@

>>>>

This section describes the approach taken to study surfactants. Here,
we always adopt the hydrid method, i.e., LB for the hydrodynmic
equations and finite difference for the associated Cahn-Hilliard
equations. Coupling between the two sectors is via the velocity
field and the external force on the fluid computed via the
divergence of the `chemical stress'.

In this model the surfactant appears as a concentration field
$\psi(\mathbf{r};t)$, alongside the compositional order parameter
$\phi(\mathbf{r};t)$ as appears in the binary fluid. There are
two corresponding Cahn-Hilliard equations: first,

\begin{equation}
\partial_t \phi + \partial_\alpha \phi u_\alpha = 
\partial_\alpha M_\phi \partial_\alpha \mu_\phi,
\end{equation}
and second
\begin{equation}
\partial_t \psi + \partial_\alpha \psi u_\alpha = 
\partial_\alpha M_\psi \partial_\alpha \mu_\psi.
\end{equation}

\subsection{Free Energy}

A number of different free energies have been used to study
surfactants [cite Yeomans]. Here, the free energy is based on
that used by van der Graff and van der Sman and others [cite].
There are a number of different contributions to the free
energy density:
\begin{equation}
f = f_\phi + f_\psi +f_1 + f_{ex}.
\end{equation}
The first is exactly that used for the binary fluid, i.e.,
\begin{equation}
f_\phi (\phi, \partial_\alpha\phi) =
{\textstyle \frac{1}{2}} A \phi^2 + {\textstyle \frac{1}{4}} B \phi^4
+ {\textstyle \frac{1}{2}} \kappa (\partial_\alpha \phi)^2.
\end{equation}
The surface tension and the interfacial width are determined from the
choice of parameters $A,B,\kappa$ as before.
The second term in the free energy density
represents the entropic
energy of mixing of the surfactant with the bulk phases
\begin{equation}
f_\psi (\psi) = kT \left[ \psi \ln\psi + (1 - \psi)\ln(1-\psi) \right]
\end{equation}
where the parameter $kT$ sets the energy scale for the whole system.
The surfactant order parameter, $\psi$, is unity for a saturated
interface and zero for a plain interface.

The surfactant lowers the energy of the interface where it is adsorbed
there, represented by the term
\begin{equation}
f_1(\psi, \partial_\alpha \phi) 
= -{\textstyle \frac{1}{2}} \epsilon \psi (\partial_\alpha \phi)^2
  -{\textstyle \frac{1}{2}} \beta  \psi^2 (\partial_\alpha \phi)^2
\end{equation}
where $\epsilon$ and $\beta$ are parameters. Note that only the
first term appears in \cite{vandergraaf}, while the second term is
based on that appearing in, e.g., \cite{diamant}. If $\beta = 0$,
the surfactant profile at the interface is symmetric (so-called
Langmuir adsoption) whereas the term in $\beta$ induces an
asymmetry (Frumpkin adsorption).
Finally,  $f_{ex}$ is [van der Sman claims]
an enthalpic contribution introduced 
to stabilise the interface \cite{theissengompper}
\begin{equation}
f_{ex} (\psi, \partial_\alpha \phi) = {\textstyle\frac{1}{2}} W \psi \phi^2,
\end{equation}
with $W$ a parameter.

\subsection{Chemical potentials}

The respective chemical potentials are obtained from the functional
derivatives of the free energy $\mu_\phi = \delta F / \delta \phi$ and
$\mu_\psi = \delta F / \delta \psi$. We obtain:
\begin{equation}
\mu_\phi =
A\phi + B\phi^3 -\kappa \partial_\alpha^2 \phi
+ W\psi\phi + \epsilon\psi\partial_\alpha^2 \phi
+ \epsilon \partial_\alpha \phi \partial_\alpha \psi
+ 2\beta \psi\partial_\alpha\phi \partial_\alpha \psi
+ \beta \psi^2 \partial_\alpha^2 \phi,
\end{equation}
and
\begin{equation}
\mu_\psi = kT \left[ \ln\psi - \ln(1-\psi) \right]
+ {\textstyle \frac{1}{2}}W\phi^2
- {\textstyle \frac{1}{2}}\epsilon (\partial_\alpha \phi)^2
- \beta\psi (\partial_\alpha \phi)^2.
\end{equation}
As before, the pressure tensor is $P_{\alpha\beta} = p_0\delta_{\alpha\beta}
+ \tilde{P}_{\alpha\beta}$ with the isotropic part being determined as (e.g.,
\cite{theissengompper})
\begin{equation}
p_0 = \phi \mu_\phi + \psi \mu_\psi - f(\phi, \psi, \partial_\alpha \phi).
\end{equation}
There is no closed form for the tensor part, which must be chosen to
comply with the constraint that in mechanical equilibrium
\begin{equation}
\partial_\alpha P_{\alpha\beta} = (\phi\partial_\alpha\mu_\phi +
\psi\partial_\alpha\mu_\psi)\delta_{\alpha\beta}.
\end{equation}
In this way we obtain for the isotropic contribution
\begin{eqnarray}
p_0 = {\textstyle \frac{1}{2}}A\phi^2 + {\textstyle \frac{3}{4}}B\phi^4 
- {\textstyle\frac{1}{2}}\kappa (\partial_\alpha\phi)^2 
- \kappa \phi \partial_\alpha^2 \phi
- kT \ln(1-\psi) + W\psi\phi^2
\nonumber\\
+ \epsilon \phi \partial_\alpha \phi \partial_\alpha \psi
+ \epsilon\phi\psi \partial_\alpha^2 \phi
+ 2 \beta\phi\psi \partial_\alpha \phi \partial_\alpha\psi
+ \beta \phi \psi^2 \partial_\alpha^2 \phi
- {\textstyle\frac{1}{2}} \beta \psi^2 (\partial_\alpha \phi)^2
\end{eqnarray}
and for the tensor part we can have
\begin{equation}
\tilde{P}_{\alpha\beta} =
(\kappa - \epsilon \psi - \beta\psi^2) 
\partial_\alpha \phi \partial_\beta \phi.
\end{equation}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
\hline
Name & $\eta$ & $-A,B$ & $\kappa$ & $kT$ & $\epsilon$ & $\beta$ & $W$
     & $M$ & $M_\psi$\\
\hline
R027a & 1/6 & 0.020833 & 0.12 & 5.6587$\times 10^{-4}$ & 0.06 & 0.0000 & 0.0 
      & 0.15 & 2.0 \\
\hline
R028a & 1/6 & 0.020833 & 0.12 & $2.82743\times 10^{-4}$ & 0.06 & 0.0000 & 0.0 
      & 0.15 & 2.0 \\
\hline
R029a & 1/6 & 0.020833 & 0.12 & 5.6587$\times 10^{-4}$ & 0.12 & 0.0000 & 0.0 
      & 0.15 & 2.0 \\
\hline
R030a & 1/6 & 0.020833 & 0.12 & $2.82743\times 10^{-4}$ & 0.12 & 0.0000 & 0.0 
      & 0.15 & 2.0 \\
\hline
\end{tabular}
\caption{Parameters for $\xi_0 = 3.39$ and $\epsilon = \kappa/2$ and
$\epsilon = \kappa$.}
\end{center}
\end{table}

@@

\subsection{Calibration of the Surface Tension}

The binary fluid model introduces an interfacial tension $\sigma$ which
can be predicted from the choice of free energy \cite{swift,kendon}.
In particular, Kendon et al. 2000 described the process of calibrating
the actual interfacial tension of the model fluid. This was based on
a measurement of the Laplace pressure difference between the fluid
inside and outside a spherical droplet via
\begin{equation}
\Delta p = 2 \sigma / R
\end{equation}
where $R$ is the radius of the spherical drop at equilibrium. This
measurement appears quite difficult in practice owing to tendency of
any initial drop to evaporate \cite{yue2007}.

\subsubsection{New measurement}

Here, we prefer a measure based on the expected profile of the
order parameter $\phi$ and the free energy density at a flat interface.
In order to sample all orientations of the interface with respect to
the lattice we again take a large droplet of fluid A initialised at
rest in fluid B. We assume that the interface is locally flat and so
the procedure holds.

If a droplet with radius $r_0$ is constructed with an initial interfacial
profile of $\phi(r) = \tanh{(r - r_0) / \xi_0}$
with $r$ in the radial direction,  then the free energy density will be
(with $-A = B$)
\begin{equation}
 e(r) = -{\scriptstyle \frac{1}{4}} |A| [1 - \mathrm{sech}^4 (r-r_0)/\xi_0]
+ {\scriptstyle \frac{1}{2}}
 (\kappa/\xi_0^2)\, \mathrm{sech}^4{(r-r_0)/\xi_0}.
\end{equation}
Obtaining a fit to $\xi_0$ from the measured interfacial profile of
$\phi (r)$ and to $ e_0 = \kappa / 2 \xi_0^2$ from the peak in the measured
profile of free energy
density allows an estimate of the interfacial tension to be made:
\begin{equation}
\sigma = 2(8/9)^{1/2} \xi_0 |e_0|.
\end{equation}
This measurement is taken after the initial profile has been allowed
to relax for at least 10 times the larger of the diffusion time
$\xi_0^2 / M$ or the momentum diffusion time $\xi_0^2 / \eta$.

We have measured the surface tension using the following approaches.

\textit{Method 1}.

LB for two distributions with relaxation of order parameter
following Kendon etal. The force on the fluid is applied
via the equilibrium stress following Swift et al.

\textit{Method 2}.

As method 1 but using the realxation of order parameter
following Stratford etal.

\textit{Method 3}.

As Method 1 but with force via divergence of the chemical stress.

\textit{Method 4}.

As Method 3 but with force via divergence of the chemical stress.

\textit{Method 5}.

Using finite difference (1st order upwind) for the order parameter.

\textit{Method 6}.

Using finite difference (3rd order upwind).

\subsubsection{Results}

We follow the parameter sets in Kendon et al Table~2. These
are reproduced here in Table~\ref{tab:r1}. Measured values for
Run028 parameters are show in Table~\ref{tab:newr028}.

\begin{table}
\begin{center}
\begin{tabular}{llllllll}
\hline
Run & $-A, B$ & $\kappa$ & $\eta$ & $M$ & $\sigma$  & $L_0$ & $T_0$ \\
\hline
Rks001 & 0.00625 & 0.004 & 1.41 & 0.05 & 0.0047 & 422 & 1.26e10$^5$\\
Rks002 & 0.00625 & 0.004 & 0.65 & 0.25 & 0.0047 & 89.7 & 1.24e10$^4$\\
Run028 & 0.083 & 0.053 & 1.41 & 0.05  & 0.0625 & 31.8 & 718 \\
Run022 & 0.0625 & 0.04 & 0.5 & 0.25   & 0.0471 & 5.31 & 56.3 \\
Run029 & 0.0625 & 0.04 & 0.2 & 0.15   & 0.0471 & 0.849 & 3.61 \\
Run020 & 0.00625 & 0.004 & 0.025 & 2.00 & 0.00471 & 0.133 & 0.704 \\
Run030 & 0.00625 & 0.004 & 0.0065 & 1.25 & 0.00471 & 0.00829 & 0.0124 \\
Run019 & 0.003125 & 0.002 & 0.0014 & 4.00 & 0.00236 & 0.000831 & 0.000493 \\
Run032 & 0.00125  & 0.0008 & 0.0005 & 5.0 & 0.000943 & 0.000265 & 0.000141\\
\hline  
\end{tabular}
\end{center}
\caption{Table showing common parameter values following Kendon et al.
with the theoretical interfacial tension, $L_0$ and $T_0$. Note that
the order parameter mobility is used here (cf $\tilde{M} = 2M$) in
Kendon et al. (KS values are extra sets used to low 1/$\dot{\gamma}T_0$.)}
\label{tab:r1}
\end{table}

\begin{table}
\begin{tabular}{lllllll}
\hline
Method & $\xi_0$ & $e_{min}$ & $e_{max}$ & $\sigma$  & $\phi_{min}$
& $\phi_{max}$\\
\hline
\multicolumn{7}{c}{Run028}\\
\hline
Theory   & 1.13 & -0.0207 &  0.0207 & 0.0625 & -1.0 & 1.0 \\
Method 1 & 1.00 & -0.0207 & 0.0128 & 0.063(3) & -1.0091 & 1.0028 \\
Method 2 & 1.00 & -0.0207 & 0.0126 & 0.063(1) & -1.0113 & 0.9993 \\
Method 3 & && & & & \\
Method 4 & && & & & \\
Method 5 & && & & & \\
Method 6 & --& --& --& --& --& --\\
\hline
\multicolumn{7}{c}{Run029}\\
\hline
Theory & 1.13 & -0.0156 & 0.0156 & 0.0471  & -1.0 & 1.0\\
Method 2 & 0.99 & -0.0156 & 0.010(?) & 0.048(4) & -1.0110 & 1.0003\\ 
Method 6 & 1.02 & -0.0156 & 0.0099 & 0.049(0) & -1.0048 & 1.0002\\
\hline
\multicolumn{7}{c}{Run030}\\
\hline
Theory & 1.13 & -0.00156 & 0.00156 & 0.00471 & -1.0 & 1.0\\
Method 2 & 1.04 & -0.00156 & 0.0008(?) & 0.0047(1) & -1.0167 & 1.0063\\
Method 6 & 1.08 & -0.00156 & 0.0007(?) & 0.0048(0) & -1.0028 & 1.0031\\
\hline
\multicolumn{7}{c}{Run019}\\
\hline
Theory & 1.13 & -0.000781 & 0.000781 & 0.00236 & -1.0 & 1.0\\
Method 2 & 1.29(?) & -0.000781 & 0.00015(?) & 0.0022(8) & -1.0312 & 1.0242\\
Method 6 & 1.23(?) & 0.0010(?) &&  0.0024(?) & -1.0085 & 1.0019\\
\hline 
\multicolumn{7}{c}{Run032}\\
\hline
Theory   & 1.13 &  -0.000312 & 0.000312 & 0.000943  & -1.0 &1.0\\
Method 2 & 1.07  & -0.000312& 0.000157 & 0.00094(7) & -1.0511 & 1.0311\\
Method 6 & 1.6(5)& -0.000312 & not st. & 0.0010(7) &-1.0056 & 1.0010\\
\hline
\end{tabular}
\caption{Results for the different methods for Run208 parameters.}
\end{table}

\subsection{User input}

\inputkey{free\_energy surfactant}

Currently under reconstruction.

This is for use with a two order parameter model the first of
which is the composition, as for the symmetric free energy,
while the second $\psi$ represents surfactant concentration.
The free energy density is made up of a number of terms:

\begin{equation}
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
\end{equation}
is the standard symmetric term related to the composition.
These parameters are represented in the input as

\inputkey{surf\_A}
\inputkey{surf\_B}
\inputkey{surf\_kappa}

\begin{equation}
D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
\end{equation}
represents the energy of surfactant in the bulk phase with surfactant
concentration $0 < \psi < 1$ with a single parameter $D$;
\begin{equation}
-{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
-{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
\end{equation}
where the term in $\epsilon$ represents the energy reduction by
adsorbing surfactant at the interface and the term in $\beta$
provides an additional favourable term modelling the fact that
the molecules like to line up; finally, there is an extra term
\begin{equation}
+{\textstyle \frac{1}{2}} W \psi \phi^2
\end{equation}
which is included to stabilise the interface \cite{theissengompper}.

The corresponding keys for the input file are:

\inputkey{D} is the bulk surfactant contribution parameter

\inputkey{epsilon} is the parameter $\epsilon$

\inputkey{W} is the parameter $W$

\inputkey{beta} is the value of the parameter $\beta$.

\inputkey{phi\_b} the initial, uniform, background surfactant
concentration $\psi_b$.

\inputkey{mobility\_psi} Sets the (uniform) mobility for the surfactant
as appropriate for the Cahn-Hilliard equation.
